---
title: "NBA Player Efficiency Rating RS 2017-18"
author: "Massimo Malandra"
date: "17/04/2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Top 100 NBA Players 2017-18 PER

This brief presentation focuses on the **TOP 100 NBA players** *per Player Efficiency Rating (PER) during the 2017-18 regular season* and illustrates how their salary figures are distributed.

The size of the points in the scatterplot represent the total games  played during the season and its color indicates in which of the two Conferences they played.

## What is PER?

As per Basketball Reference (www.basketball-reference.com)'s definition:


- The *Player Efficiency Rating (PER)* is a per-minute rating developed by ESPN.com columnist **John Hollinger**. 
In John's words, "The PER sums up all a player's positive accomplishments, subtracts the negative accomplishments, and returns a per-minute rating of a player's performance."


More info about PER:
<br>
https://www.basketball-reference.com/about/per.html
<br>
http://www.espn.com/nba/columns/story?columnist=hollinger_john&id=2850240


```{r, include = FALSE}
library(xlsx)
library(readr)
library(ggplot2)
library(plotly)
library(dplyr)


stats <- read.xlsx("nbastats.xlsx", 
                sheetIndex = 1)

contracts <- read.xlsx("contracts.xlsx", 
                   sheetIndex = 1)

teams <- read.xlsx("teams.xlsx", 
                       sheetIndex = 1)

# RK (ranking) column has few 'NA':
stats$RK <- 1:100

# Use substr function combined with regexpr:
# the regexpr function returns the position in which the comma
# is found in the Player columns. Subtracting -1 to that, return
# the position BEFORE the comma is found, hence the position
# of the last character of the player's name. I use that as the
# 'stop' parameter of the substr function.
stats$names <- substr(stats$PLAYER, 1, regexpr(",", stats$PLAYER)-1)
stats$team <- substr(stats$PLAYER, regexpr(",", stats$PLAYER)+1, length(stats$PLAYER)) 

names(contracts)[names(contracts) == "Player"] <- "names"

# fixing few differences in names typing:

stats$names <- replace(stats$names, stats$names == "CJ McCollum", 
                       "C.J. McCollum") 
stats$names <- replace(stats$names, stats$names == "Larry Nance Jr.", 
                       "Larry Nance") 
stats$names <- replace(stats$names, stats$names == "Otto Porter Jr.", 
                       "Otto Porter") 
stats$names <- replace(stats$names, stats$names == "TJ Warren", 
                       "T.J. Warren") 

# # I will not use the substr until the length of the full string (length(merged$team)),
# # because I will also solve the problem of mutliple teams for one 
# # player, considering only the  first team
# merged$team1 <- substr(merged$team, 2, length(merged$team))
merged <- merge(x = stats, 
                y = contracts[ , c("names", "X2017.18", "Tm")], 
                by = "names", all.x = TRUE)


# Check if there are missing in RK:
merged[!complete.cases(merged$RK), ]


# Removing Greg Monroe who appears twice
merged <- merged[!(merged$names == "Greg Monroe" & 
               merged$X2017.18 == "$5,000,000"), ]

str(merged)


merged <- merge(x = merged, y = teams[ , c("team", "conference")],
                 by.x = "Tm",
                 by.y = "team",
                 all.x = TRUE)

# Check if there are missing values:
merged[!complete.cases(merged$conference), ]


# Changing X2017.18 var name:
names(merged)[names(merged) == "X2017.18"] <- "salary"

# and type, using the parse function in readr:
merged$salary <- parse_number(merged$salary)

# Converting PER from factor to numeric:
merged$PER <- as.numeric(levels(merged$PER)[merged$PER])

merged$salaryMln <- round(merged$salary / 1000000, 2)
```


***


```{r, echo = FALSE, fig.align = 'center'}
plot_ly(merged, x = merged$PER, y = merged$salaryMln, 
        type = "scatter", mode = "markers", 
        size = merged$GP, alpha = 0.6,
        color = merged$conference,
        colors = c("red", "blue"),
        text = ~paste(merged$names, merged$Tm)) %>%
  layout(title = "Top 100 NBA Players per Player Efficiency Rating (PER)",
         titlefont = list(size = 16, 
                          color = "blue"),
    xaxis = list(title = "Player Efficiency Rating (PER)",
                 nticks = 10,
                 tickfont = list(size = 11),
                 titlefont = list(size = 11)),
    yaxis = list(title = "Salary (Millions of Dollar)",
                 nticks = 10,
                 tickfont = list(size = 11),
                 titlefont = list(size = 11)))
    
```


## Some quick insight:

The most efficient player at the lowest salary is *Karl-Anthony Towns* from Minnesota Timberwolves, who had an efficiency rating of 24.99 during the regular season - very close to players like Russell Westbrook, LaMarcus Aldridge, Kyrie Irving - having a salary more than 3 times lower than the ones of its All-Star colleagues (6.22 Mln).

Montrezl Harrell (LAC), Clint Capela (HOU), Nikola Jokic (DEN) closely follow.

